---
title: "DS311 - R Lab Assignment"
author: "Giovanni Vallecillo"
date: "`(4/29/25)`"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Assignment 1

-   In this assignment, we are going to apply some of the build in data set in R for descriptive statistics analysis.
-   To earn full grade in this assignment, students need to complete the coding tasks for each question to get the result.
-   After finished all the questions, knit the document into HTML format for submission.

### Question 1

Using the **mtcars** data set in R, please answer the following questions.

```{r}
# Loading the data
data(mtcars)

# Head of the data set
head(mtcars)
```

a.  Report the number of variables and observations in the data set.

```{r}
# Enter your code here!
variable_number <- ncol(mtcars)
observation_number <- nrow(mtcars)

# Answer:
print(paste("There are total of", variable_number , "variables and", observation_number , "observations in this data set."))
```

b.  Print the summary statistics of the data set and report how many discrete and continuous variables are in the data set.

```{r}
# Enter your code here!
summary(mtcars)
str((mtcars))
# Answer:
print("There are 5 discrete variables and 6 continuous variables in this data set.")
```

c.  Calculate the mean, variance, and standard deviation for the variable **mpg** and assign them into variable names m, v, and s. Report the results in the print statement.

```{r}
# Enter your code here!
mpg_mean <- mean(mtcars$mpg)
mpg_variance <- var(mtcars$mpg)
mpg_sd <- sd(mtcars$mpg)

mpg_mean
mpg_variance
mpg_sd
print(paste("The average of Mile Per Gallon from this data set is ",round(mpg_mean,3)  , " with variance ",round(mpg_variance,3)  , " and standard deviation",round(mpg_sd,3)  , "."))
```

d.  Create two tables to summarize 1) average mpg for each cylinder class and 2) the standard deviation of mpg for each gear class.

```{r}
# Enter your code here!
average_mpg_per_cyl <- aggregate(mpg~cyl, data=mtcars, FUN=mean)
print(average_mpg_per_cyl)

average_mpg_per_gear <- aggregate(mpg~gear, data=mtcars, FUN=sd)
print(average_mpg_per_gear)

```

e.  Create a crosstab that shows the number of observations belong to each cylinder and gear class combinations. The table should show how many observations given the car has 4 cylinders with 3 gears, 4 cylinders with 4 gears, etc. Report which combination is recorded in this data set and how many observations for this type of car.

```{r}
# Enter your code here!
mtcars_crosstab <- table(mtcars$cyl,mtcars$gear)
print(mtcars_crosstab)

print("The most common car type in this data set is a car with 8 cylinders and 3 gears. There are total of 12 cars belong to this specification in the data set.")
```

------------------------------------------------------------------------

### Question 2

Use different visualization tools to summarize the data sets in this question.

a.  Using the **PlantGrowth** data set, visualize and compare the weight of the plant in the three separated group. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your findings.

```{r}
# Load the data set
data("PlantGrowth")

# Head of the data set
head(PlantGrowth)

# Enter your code here!
boxplot(weight~group,
                data = PlantGrowth, 
                main = "Plant Group Weights",
                xlab = "Plant Group",
                ylab = "Plant Weight",
                col = "lightgreen")

```

Result:

=\> Report a paragraph to summarize your findings from the plot!

b.  Using the **mtcars** data set, plot the histogram for the column **mpg** with 10 breaks. Give labels to the title, x-axis, and y-axis on the graph. Report the most observed mpg class from the data set.

```{r}
hist(mtcars$mpg,
     breaks = 10,
     main = "Miles per Gallon Observations",
     xlab = "Miles per Gallon (MPG)",
     ylab = "Frequencies",
     col = "lightgrey")


print("Most of the cars in this data set are in the class of 15 miles per gallon.")
```

c.  Using the **USArrests** data set, create a pairs plot to display the correlations between the variables in the data set. Plot the scatter plot with **Murder** and **Assault**. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your results from both plots.

```{r}
# Load the data set
data("USArrests")

# Head of the data set
head(USArrests)

# Enter your code here!
pairs(USArrests,
      main = "Pairs Plot of US Arrests Data")

# US Arrests for Murder & Assault Rates Table
plot(USArrests$Murder, USArrests$Assault,
     main = "Assault vs. Murder Rates by State",
     xlab = "Assault rate",
     ylab = "Murder rate",
     pch = 20, col = "black")

```

Result:

=\> Report a paragraph to summarize your findings from the plot! The pairs plot details extensive positive relation between Murder, Assault, and Rape rates across the United States. the scatterplot of Assault vs. Murder in particular reveals the strong positive trends between states that experience high rates of Assualt and Murder leading to a visualization that these those rates have great correlation in those states they are reported.

------------------------------------------------------------------------

### Question 3

Download the housing data set from www.jaredlander.com and find out what explains the housing prices in New York City.

Note: Check your working directory to make sure that you can download the data into the data folder.

```{r, echo=FALSE}
# Load and clean the housing data set 
download.file(url='https://www.jaredlander.com/data/housing.csv',
              destfile='housing.csv', mode='wb')
housingData <- read.csv('housing.csv')
housingData <- subset(housingData, 
                      select = c("Neighborhood", "Market.Value.per.SqFt", "Boro", "Year.Built"))
housingData <- na.omit(housingData)                   
```

a.  Create your own descriptive statistics and aggregation tables to summarize the data set and find any meaningful results between different variables in the data set.

```{r}
# Head of the cleaned data set
head(housingData)

# Enter your code here!
aggregate(Market.Value.per.SqFt~Boro, data = housingData, FUN = mean)

aggregate(Market.Value.per.SqFt~Year.Built, data = housingData, FUN = mean)

aggregate(Market.Value.per.SqFt~Neighborhood, data = housingData, FUN = mean)


```

b.  Create multiple plots to demonstrates the correlations between different variables. Remember to label all axes and give title to each graph.

```{r}
# Enter your code here!
plot(housingData$Market.Value.per.SqFt, housingData$Year.Built,
     main = "Market Value per Square Foot vs. Year Built",
     xlab = "Market Value per Square Foot",
     ylab = "Year Built",
     pch = 20, col = "purple")

boxplot(Market.Value.per.SqFt~Boro, data = housingData,
         main = "Market Value per Square Foot by Boro",
         xlab = "Boro",
         ylab = "Market Value per Square Foot",
         col = "orange")


```

c.  Write a summary about your findings from this exercise.

=\> Enter your answer here! A summary of the data and the outlined boxplots/plots give insight into the averge market value for a New York real estate in relation to when they were built and how large they are. There is a strong positive correlation between market value real estate has and the square footage of the real estate Manhattan appears to have the largest concentration of higher valued real estate on average based on the value per square foot. There is also more positive correlation between the market value based on square footage and the year the home was built. Older real estate built before 1940-1960 on average possess lower market value per square foot which gives a glimpse into the historical aspect that New York's real estate market operates on.
